---
title: "Should You Wait? Solving a Niche NYC Problem"
author: "Ravi Brenner"
date: "2026-03-02"
categories: [projects]
execute: 
  warning: false
  message: false
format:
  html:
    resources: 
      - shinylive-sw.js
---

Here is a common problem I face: I live near a subway stop in New York City that has both express and local trains. If they arrive at the same time, it's much faster to take the express train, which stops less frequently, than the local train, which makes every single stop.

The Metropolitan Transit Authority (MTA) helpfully displays the estimated time until the next train arrives at the station. Sometimes, I am trying to get from my stop at 96th Street to 34th Street. The local #1 train is waiting at the platform, about to leave, and the express #3 train is 3 minutes away. Should I wait?

Fortunately, the MTA [publishes some data](https://www.mta.info/developers) that can help with this. Their exact transit schedule is published and available as a [GTFS (General Transit Feed Specification)](https://gtfs.org/) feed. Basically, this is an easily accessible set of data tables that specifies the normal operating schedule of the entire NYC transit system.

So I figured I would solve my problem once and for all. Should you wait for the express train or just get on the local? Of course it depends on your start and end points. I also am assuming that everything is operating as normal, so I will not be taking into account construction or real-time delays. Finally, I am restricting this analysis to express stops where there are two possible trains. Of course in real life you might have a more complicated route, and it could even pay to switch on and off the express train between local endpoints if the timing is right. But ultimately it boils down to whether the fastest route between two express stations involves waiting for an express train, or taking a sooner-to-arrive local train.

I will be walking through the R code I used to do this. If you just want answers, skip down below.

# Getting the data

Getting the data from MTA is fairly straightforward using the [tidytransit](https://r-transit.github.io/tidytransit/) R package.

```{r}
library(tidyverse)
library(tidytransit)
library(gtfsrouter)
library(sf)
library(ggspatial)

mta <- read_gtfs("https://rrgtfsfeeds.s3.amazonaws.com/gtfs_subway.zip")
```

The MTA subway data is a list of tables, containing various data on all the train routes and stops.

```{r}
names(mta)
```

The first thing we need to do is identify which stops are express stops. This can be quite complicated, especially since sometimes express trains run at specific times of day, and routes can change on the weekend. To start, let's find, for each subway stop, what the next stop is for that trip. We will also filter for weekday, morning (9am-12pm) services.

```{r}
adjacent_stops <- 
  # Get weekday trips
  mta$trips |>
  filter(service_id == "Weekday") |>
  # Join with the stop times for these trips
  inner_join(mta$stop_times, by = join_by("trip_id")) |>
  filter(arrival_time >= hms("08:00:00") & arrival_time <= hms("12:00:00")) |>
  # For each trip, get the stops and stop sequence
  select(route_id,trip_id, stop_id, stop_sequence) |>
  group_by(trip_id) |>
  arrange(stop_sequence) |>
  mutate(next_stop_id = lead(stop_id)) |>
  filter(!is.na(next_stop_id)) |>
  ungroup() |>
  distinct(route_id,stop_id, next_stop_id)
```

We now have a table of adjacent weekday stops, forming the subway network. What does this look like? We can join these adjacent stops with the `stops` table to get their geographic locations, and then map it.

```{r}
# Get the lattitude and longitude for each stop
adj_stops_loc <- adjacent_stops |>
  inner_join(mta$stops, by = join_by("stop_id" == "stop_id")) |>
  inner_join(mta$stops, by = join_by("next_stop_id" == "stop_id"),
             suffix = c("_orig","_dest")) |>
  # add the route color
  inner_join(mta$routes |>
               select(route_id, route_color), by = "route_id") |>
  mutate(route_color = str_c("#",route_color))

# Turn the origin-destination pairs into lines
network_lines <- adj_stops_loc |>
  rowwise() |>
  mutate(geometry = st_sfc(
    st_linestring(
      matrix(c(
        stop_lon_orig, stop_lat_orig,
        stop_lon_dest, stop_lat_dest
      ), ncol = 2, byrow = TRUE)), 
    crs = 4326)) |>
  st_as_sf()

# Create a separate object for the stops (points)
network_points <- mta$stops |>
  st_as_sf(coords = c("stop_lon", "stop_lat"), crs = 4326)

ggplot() +
  annotation_map_tile("cartolight",zoom = 11) +
  # Draw the subway lines
  geom_sf(data = network_lines, 
          aes(color = route_color),
          size = 0.5, alpha = 0.7) +
  # Draw the stations
  geom_sf(data = network_points, size = 0.2, color = "black") +
  theme_void() +
  labs(title = "NYC Subway System",
       subtitle = "Weekdays 8am-12pm") +
  scale_color_identity()
```

Great! We have successfully generated a map that looks a lot like the [NYC subway map](https://www.mta.info/map/5256). It's not an exact match, but its close.

Next, we can identify express stops. An express stop should have several characteristics:

- The origin station should have multiple destinations (a local and an express destination)

- There should be multiple routes that stop at the station (so that you could have the option to choose)

  - A destination station should have multiple origins, at least 1 of which should be an express origin

  - A destination station should also have multiple routes that stop at the station

- There should be at least 2 trains at the origin station that are also at the destination station. We can do this by looking at the intersection of the routes that stop at each origin-destination station pair.

```{r}
# Start with the origin stations
orig_stops_routes <- adj_stops_loc |>
  group_by(stop_id) |>
  summarize(num_destinations = n_distinct(next_stop_id),
            dest_list = list(unique(next_stop_id)),
            num_routes_orig = n_distinct(route_id),
            route_list_orig = list(unique(route_id))) |>
# They should all have at least 2 destinations and 2 routes to be and express station
  filter(num_destinations > 1,
         num_routes_orig > 1)

# Destination stations
dest_stops_routes <- adj_stops_loc |>
  group_by(next_stop_id) |>
  summarize(num_origins = n_distinct(stop_id),
            orig_list = list(unique(stop_id)),
            num_routes_dest = n_distinct(route_id),
            route_list_dest= list(unique(route_id))) |>
# Should all have 2 origins and 2 routes to be an express destination
  filter(num_origins > 1,
         num_routes_dest > 1)

express_stations_full <- adj_stops_loc |>
# Recreate those same columns basically
  select(-route_color) |>
  group_by(stop_id) |>
  mutate(num_destination = n_distinct(next_stop_id),
         dest_list = list(unique(next_stop_id)),
         num_routes_orig = n_distinct(route_id),
         route_list_orig = map(list(unique(route_id)),sort)) |>
  group_by(next_stop_id) |>
  mutate(num_origins = n_distinct(stop_id),
         orig_list = list(unique(stop_id)),
         num_routes_dest = n_distinct(route_id),
         route_list_dest= map(list(unique(route_id)),sort)) |>
  ungroup() |>
# Filter only the origins and destinations from above
  filter(stop_id %in% orig_stops_routes$stop_id,
         next_stop_id %in% dest_stops_routes$next_stop_id) |>
  mutate(routes_intersect = map2(route_list_orig, route_list_dest, intersect),
         num_routes_intersect = map_dbl(routes_intersect, length)) |>
# The origin an destination should have at least 2 overlapping train lines too
  filter(num_routes_intersect > 1) |>
  select(-route_id) |>
  distinct()

express_stations <-
  express_stations_full |>
  ungroup() |>
  select(parent_station_orig) |>
  distinct()
```

Now that we have our express stations we can take a quick look at them on a map again

```{r}
# Create a separate object for the stops (points)
express_points <- express_stations_full |>
  st_as_sf(coords = c("stop_lon_orig", "stop_lat_orig"), crs = 4326)

ggplot() +
  annotation_map_tile("cartolight",zoom = 11) +
  # Draw the subway lines
  geom_sf(data = network_lines, 
          aes(color = route_color),
          size = 0.5, alpha = 0.7) +
  # Draw the stations
  geom_sf(data = express_points,
          size = 2, color = "black") +
  theme_void() +
  labs(title = "NYC Subway System",
       subtitle = "Weekdays 8am-12pm") +
  scale_color_identity()
```

Now bad! This probably isn't a perfect list of express stops, but I think it's a pretty good list. I tried to verify it manually by referring to the real subway map, although I may have accidentally included some stops or missed others.

Finally, we can calculate a time between stations on the same lines.

```{r}
# First get the general arrival times of stations relative to the origin stop
travel_times <- mta$stop_times |>
  inner_join(
    mta$trips |>
      select(trip_id, route_id, service_id, direction_id)
  ) |>
  filter(
    service_id == "Weekday",
    arrival_time >= hms("08:00:00") & arrival_time <= hms("12:00:00")
  ) |>
  group_by(trip_id) |>
  # Make sure the first stop is present, for timing purposes
  filter(any(stop_sequence == 1)) |>
  # Calculate avg time of departure, and relative time of arrival at station
  mutate(
    start_time = min(departure_time),
    station_time = arrival_time - start_time
  ) |>
  group_by(route_id, direction_id, stop_id) |>
  # get average and number of trips
  summarize(avg_station_time = mean(station_time), num_trips = n()) |>
  mutate(direction_letter = str_sub(stop_id, -1, -1)) |>
  inner_join(mta$stops) |>
  ungroup()
```

Now, join each express travel time to every other stop with the same route and direction.

```{r}
travel_time_grid <- travel_times |>
  # Select columns and join to itself so that we get all origin and destination pairs
  select(
    route_id,
    direction_letter,
    stop_id,
    stop_name,
    avg_station_time,
    num_trips
  ) |>
  full_join(
    travel_times |>
      select(route_id, direction_letter, stop_id, stop_name, avg_station_time),
    by = c("route_id", "direction_letter"),
    relationship = "many-to-many",
    suffix = c("_orig", "_dest")
  ) |>
  # make sure destination is after origin
  filter(avg_station_time_dest > avg_station_time_orig) |>
  # filter so that we only get express stops
  filter(
    stop_id_orig %in% express_stations_full$stop_id,
    stop_id_dest %in% express_stations_full$next_stop_id
  ) |>
  # calculate lag times
  mutate(
    lag_time = hms::hms(as.numeric(
      avg_station_time_dest - avg_station_time_orig
    ))
  ) |>
  group_by(stop_id_orig, stop_id_dest) |>
  # enforce minimum number of trips for quality
  filter(num_trips > 3) |>
  # some better route list formatting
  mutate(
    route_list = list(unique(route_id)),
    route_list = map_chr(route_list, \(x) str_c(sort(x), collapse = "/")),
    full_name_orig = str_c(stop_name_orig, " ", route_list)
  ) |>
  # ensure that there is >1 route stopping
  filter(str_length(route_list) > 1) |>
  # get route colors
  inner_join(
    mta$routes |>
      select(route_id, route_color),
    by = "route_id"
  ) |>
  mutate(route_color = str_c("#", route_color)) |>
  ungroup() |>
  # cut stops where the absolute time difference is less than a minute and there are only 2 trains
  mutate(
    .by = c(stop_id_orig, stop_id_dest, route_list),
    keep = case_when(
      str_length(route_list) == 3 &
        abs(lag_time - lag(lag_time)) < hms("00:00:59") ~ 0,
      str_length(route_list) == 3 &
        abs(lag_time - lead(lag_time)) < hms("00:00:59") ~ 0,
      .default = 1
    )
  ) |>
  group_by(stop_name_orig, route_list, stop_name_dest) |>
  arrange(lag_time) |>
  mutate(wait_time = round(as.numeric(last(lag_time) - lag_time) / 60)) |>
  filter(any(wait_time > 0))

write_csv(travel_time_grid, "travel_time_grid.csv")
```

Finally, can we visualize these results at all? To keep things simple, we will focus on a single train line and group of stops, the 1/2/3 going southbound.

```{r}
travel_times |>
   inner_join(mta$routes |>
               select(route_id, route_color),
             by = "route_id") |>
  mutate(route_color = str_c("#",route_color)) |>
  ungroup() |>
  filter(
    stop_id %in%
      travel_time_grid$stop_id_orig |
      stop_id %in% travel_time_grid$stop_id_dest
  ) |>
  group_by(stop_id) |>
  mutate(
    route_list = list(unique(route_id)),
    route_list = map_chr(route_list, \(x) str_c(sort(x), collapse = "/"))
  ) |>
  filter(direction_letter == "S", route_list == "1/2/3") |>
  select(route_id, stop_id,direction_letter, route_list, stop_name, avg_station_time,route_color) |>
  group_by(route_list,route_id) |>
  arrange(avg_station_time) |>
  mutate(
    wait_time = as.numeric(avg_station_time - first(avg_station_time)) / 60
  ) |> 
  ungroup() |>
  ggplot(aes(
    x = reorder(stop_name, wait_time),
    y = wait_time,
    color = I(route_color),
    shape = route_id,
    group = route_id
  )) +
  geom_point(alpha = 0.5) +
  geom_line(alpha = 0.5) +
  labs(x = "Stop",
  y = "Average time from first stop",
  shape = "Train line",
  title = "1/2/3 Trains going south") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

You can see that, starting from the northmost station at 96 St, the gap gets wider and wider as you go south. The 2 and 3 trains are both express and are nearly overlapping. If you are traveling between intermediate stations, you would really be interested in the *difference between those gaps*, not the gaps themselves.

Here's another way of looking at the same data:

```{r}
travel_time_grid |>
  filter(direction_letter == "S", route_list == "1/2/3") |>
  mutate(lag_time = as.numeric(lag_time) / 60) |>
  ggplot(aes(
    x = reorder(stop_name_orig, -lag_time),
    y = reorder(stop_name_dest, -lag_time),
    fill = lag_time,
    label = str_c(round(lag_time), " min")
  )) +
  geom_tile() +
  geom_text(color = "white", size =3) +
  facet_wrap(~route_id) +
  theme_minimal() +
  labs(
    x = "Origin station",
    y = "Destination station",
    title = "Travel times for 1/2/3 Trains going south",
    fill = "Travel time"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

You can see here how the travel times between pairs of stations vary. Stating from 72 St and going to 14 St for example, the 1 train takes 12 minutes while the 2 and 3 trains take 8 minutes.

# Should you wait?

I have compiled the data here into a convenient tool so that you can look at a given stop, the train line, and destination, and see how long you should wait for the express. Rather than give you the travel time, the tool simply tells you, if the local train is already at the station, what is the maximum amount of time you should wait for the express train in order to arrive at your destination sooner.
```{shinylive-r}
#| standalone: true
#| viewerHeight: 600
#| resources: 
#|   - travel_time_grid.csv

library(shiny)
library(bslib)
library(dplyr)
library(ggplot2)

travel_time_grid <- readr::read_csv("travel_time_grid.csv")

ui <- page_sidebar(
  sidebar = sidebar(open = "open",
    uiOutput("orig_select"),
    uiOutput("train_select"),
    uiOutput("dest_select")
  ),
  card(plotOutput("plot"),
       max_height = "400px")
)

server <- function(input, output, session) {
  output$orig_select <- renderUI({
    origin_station_list <- travel_time_grid |>
      pull(stop_name_orig) |> 
      unique() |>
      sort()
    
    selectInput("orig_select_input","Select Origin",
                choices = origin_station_list)
  })
  
  output$train_select <- renderUI({
    req(input$orig_select_input)
    
    train_options <- travel_time_grid |>
      filter(stop_name_orig == input$orig_select_input) |>
      pull(route_list) |>
      unique() |>
      sort() 
    
    selectInput("train_select_input","Select trains",
                choices = train_options)
  })
  
  output$dest_select <- renderUI({
    req(input$orig_select_input, input$train_select_input)
    
    dest_station_list <- travel_time_grid |>
      filter(stop_name_orig == input$orig_select_input,
             route_list == input$train_select_input) |>
      pull(stop_name_dest) |>
      unique() |>
      sort()
    
    selectInput("dest_select_input","Select Destination",
                choices = dest_station_list)
  })
  
  output$plot <- renderPlot({
    req(input$orig_select_input, input$dest_select_input)
    
    df <- travel_time_grid |>
      ungroup() |>
      filter(stop_name_orig == input$orig_select_input,
             route_list == input$train_select_input,
             stop_name_dest == input$dest_select_input) |>
      select(route_id, direction_letter, lag_time, route_color) |>
      arrange(lag_time) |>
      mutate(wait_time = last(lag_time) - lag_time) |>
      mutate(wait_time_minutes = round(as.numeric(wait_time)/60),
             wait_time_minutes = case_when(wait_time_minutes == 0 ~ "If this train is at the station...",
                                           .default = paste0("Wait no more than ", wait_time_minutes," minutes"))) 
    
    df |>
      ggplot(aes(x = 1, y = reorder(route_id, -wait_time))) +
      geom_point(aes(color = I(route_color)), size = 30) +
      geom_text(aes(label = route_id),
                color = "white", 
                fontface = "bold", size = 15) +
      geom_text(aes(x = 1.7, label = wait_time_minutes), 
                hjust = 1, size = 5, fontface = "bold", family = "sans") +
      scale_x_continuous(limits = c(0.8, 2)) +
      theme_void()
  })
}

shinyApp(ui = ui, server = server)

## file: travel_time_grid.csv
{{< include travel_time_grid.csv >}}
```

Note that this does not take into account any more complicated routing, like switching train lines entirely. This also implicitly assumes that, for any truly adjacent stops (like 42 St and 34 St), you should always just take the next arriving train no matter what.

# Reflection

I had the idea for this analysis a while ago, but I was intimidated by the GTFS data format. It is tricky to work with but I've learned a lot by doing so. Kudos to the authors of the [`tidytransit`](https://r-transit.github.io/tidytransit/) package for making it relatively painless.

A major challenge here was just defining what an express station even is. It is not defined clearly in the GTFS specification from what I could tell. Moreover, I could not find any resources on the internet that took any sort of similar approach (if you know of one, please reach out!). Because of that, I found that AI models did a very poor job when I tried to ask them for help. They would give approximate solutions, or suggest fixes that would sort of work for some stations but not others. Being able to think clearly about a problem like that is still important, as AI models will always be somewhat limited by their training data.

I considered using a network analysis approach to solving this problem. The subway system is a network after all: nodes (stations) connected via directed edges (train routes). I think that would be a viable and interesting approach here too. You might need a slightly different definition of an express station, and you would have to make significant use of edge weights/values (for travel times). 